---
title: "Introgression"
output: 
  rmarkdown::html_document:
    css: style.css
vignette: >
  %\VignetteIndexEntry{pcadapt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = TRUE, include = FALSE, eval=FALSE}
require(pcadapt)
require(data.table)
require(knitr)
require(ggplot2)
opts_knit$set(root.dir = "~/Documents/thesis/git/Introgression/populus/Original_data_set_ch6_12_15/")
```

##### Download the populus dataset here and set the working directory to where the data have been downloaded:

https://www.dropbox.com/sh/85cgelc1af1s7az/AABvFT5W7NrDd_cXQHTff2Pma?dl=0


```{r, eval=FALSE}
setwd("~/Documents/thesis/git/Introgression/populus/Original_data_set_ch6_12_15/")
filename <- "populus3pops.pcadapt" 
popfile <- "populus3pops.pop" 
pop <- as.character(read.table(popfile)[, 1]) 
```

```{r, include=FALSE, eval=FALSE}
filename <- "populus3pops.pcadapt" 
popfile <- "populus3pops.pop" 
pop <- as.character(read.table(popfile)[, 1]) 
```

## Getting a complete dataset

Our method requires complete datasets, in case your data contain missing values, the package provides a function to impute missing values in two different ways, depending on whether individual labels are known or not.
In our example, `pop` contains the population information, i.e `pop[i]` returns the name of the population for which the `i`-th individual belongs to. 

```{r, eval=FALSE}
head(pop)
```

```{r, eval=FALSE}
geno <- impute.pcadapt(filename, pop = pop)$x
```

If this information is lacking, leave the `pop` argument to blank.

**N.B:** Once it is done, we recommend to save the imputed matrix in an external file so you do not need to go through the imputation step every time you open a new `R` session: 
```{r, eval=FALSE}
write.table(geno, "imputed.pcadapt", col.names = FALSE, row.names = FALSE)
```

```{r, include=FALSE, eval=FALSE}
geno <- as.matrix(fread("imputed.pcadapt"))
```

## Running the scan

Provide the function `scan.intro` with the following arguments:

- **input** a genotype matrix or a character string specifying the name of the file to be imputed.
- **K** a vector of integers specifying the components along which local ancestries may vary.
- **pop** a vector of integers or strings specifying which subpopulation the individuals belong to.
- **ancstrl.1** a string specifying the label of the ancestral population genetically closer to the hybrid population.
- **ancstrl.2** a string specifying the label of the ancestral population genetically further from the hybrid population.
- **admxd** a string specifying the label of the hybrid population.
- **ploidy** an integer specifying the ploidy of the individuals.
- **window.size** an integer specifying the window size.
- **impute** a logical value indicating whether the genotype matrix has missing values or not. 

```{r, eval=FALSE}
stat <- scan.intro( input = geno, 
                    K = 1, 
                    pop = pop, 
                    ancstrl.1 = "Trichocarpa", 
                    ancstrl.2 = "Balsamifera", 
                    admxd = "Hybrid", 
                    min.maf = 0.05,
                    ploidy = 2, 
                    window.size = 15000,
                    impute = FALSE)
```

## Analyzing the output

High peaks correspond to excess of local ancestry from `ancstrl.2`.

```{r, fig.width = 7, fig.height = 5, fig.align = 'center', eval=FALSE}
plot(stat, by.step = 10, hline = 3.0)
```