---
title: "FDR"
output: 
  rmarkdown::html_document:
    css: style.css
vignette: >
  %\VignetteIndexEntry{pcadapt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

## Description

This vignette describes a way to compare different methods, based on the 
rankings provided by the statistics they return. One way to do this is to draw 
the power as a function of the false discovery rate, which requires to compute 
the power and the false discovery rate incrementally for each method. Here is a 
quick example of how to achieve this type of comparison.

## Get two sets of statistics

Let's compare the Mahalanobis distance and the communality statistics by running
`pcadapt` on `geno3pops`.


```{r}
library(pcadapt)
file <- read.pcadapt("geno3pops", type = "example")
geno <- as.matrix(read.table(file))
gt <- 1:150
x <- pcadapt(geno, K = 2, method = "mahalanobis")
y <- pcadapt(geno, K = 2, method = "communality")
rank.stat.1 <- sort(x$pvalues, decreasing = FALSE, index.return = TRUE)$ix #any set of p-values
rank.stat.2 <- sort(y$pvalues, decreasing = FALSE, index.return = TRUE)$ix
```

## FDR and power

The functions `compute.fdr` and `compute.power` defined below respectively compute the values of FDR and power for a list of candidates (`list`) and a given list of true positives (`ground.truth`).

```{r}
compute.fdr = function(list, ground.truth){
  if (length(list) == 0){
    x <- 0
  } else {
    x <- sum(!(list %in% ground.truth)) / length(list)
  }
  return(x)
}

compute.power = function(list, ground.truth){
  if (length(ground.truth) == 0){
    warning("The list of true positives is empty.")
  } else {
    x <- sum(list %in% ground.truth) / length(ground.truth)
  }
  return(x)
}
```

Now if you want to plot the power as a function of the false discovery rate, you 
need to compute both power and FDR for any list of candidates which length 
should ideally vary from one to the total number of markers. What you want to do 
is ranking your markers according to their p-value, then compute both FDR and 
power for the marker with the lowest p-value, store both values in a data frame, 
then add the second marker with the lowest p-value, compute both FDR and power, 
store them in the data frame, and so on. The function below takes on these 
computations.

```{r}
create.fdr.pow = function(list, ground.truth, lmax = length(list), smooth = TRUE){
  fdr <- 0
  pow <- 0
  s <- seq(1, lmax, by = 1)
  for (k in s){
    l <- list[1:k]
    fdr <- c(fdr, compute.fdr(l, ground.truth))
    pow <- c(pow, compute.power(l, ground.truth))
  }
  res <- NULL
  idx <- sort(fdr, decreasing = FALSE, index.return = TRUE)$ix
  fdr <- fdr[idx]
  pow <- pow[idx]
  if (smooth == FALSE){
    res$fdr <- fdr
    res$pow <- pow
  } else if (smooth == TRUE){
    pow.old <- 0
    fdr.old <- 0
    aux.fdr <- 0
    aux.pow <- 0
    for (val in 1:length(fdr)){
      if (pow[val] >= pow.old){
        aux.fdr <- c(aux.fdr, fdr[val])
        aux.pow <- c(aux.pow, pow[val])
        pow.old <- pow[val]
      }
    }
    res$fdr <- aux.fdr
    res$pow <- aux.pow
  }
  return(res)
}
```

## Build the data frame

Now all we need is a function to generate a data frame so we can visualize the 
result with `ggplot2`.

```{r}
create.df = function(res, soft.name){
  ncol <- length(res$fdr)
  df <- array(0, dim = c(ncol, 3))
  df[, 1] <- rep(soft.name, ncol)
  df[, 2] <- res$fdr
  df[, 3] <- res$pow
  return(as.data.frame(df))
}

# Generate fdr and power values for each method:
res.1 <- create.fdr.pow(rank.stat.1, gt, smooth = TRUE)
df.1 <- create.df(res.1, "Mahalanobis")
res.2 <- create.fdr.pow(rank.stat.2, gt, smooth = TRUE)
df.2 <- create.df(res.2, "Communality")

df <- rbind(df.1, df.2)
colnames(df) <- c("Software", "FDR", "Power")
df[, 1] <- as.character(df[, 1])
df[, 2] <- as.numeric(as.character(df[, 2]))
df[, 3] <- as.numeric(as.character(df[, 3]))
```

## Display the curves

```{r, fig.width = 7, fig.height = 5, fig.align = 'center'}
require(ggplot2)
p0 <- ggplot(data = df, aes(x = FDR, y = Power)) + 
  geom_line(aes(color = Software), size = 2, na.rm = TRUE) +
  xlim(0, 1) + ylim(0, 1) +
  theme_bw() +
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 15, face = "bold"),
        title = element_text(size = 15, face = "bold"),
        legend.text = element_text(size = 15),
        legend.key.height = unit(1, "line"),
        legend.key.width = unit(3, "line")
  )
print(p0)
```

